Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-07-23 Manuscript_Nater_etal.qmd
Addition to discussion

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018]. Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). While estimation of latent parameters in missing areas may be possible (Perry de Valpine, personal communication; @schaub2021 chapter 19), this also provides an opportunity for inclusion of additional data. The line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.

>>>>

The second (and perhaps most attractive) aspect of our modelling framework in the context of future work is its spatio-temporal hierarchical structure. While we included spatial, temporal, and residual variation in our framework here, we treated them as independent. Alternatively, spatial (and temporal) correlations among parameters can be modelled explicitly, something that is commonly done e.g. in modern species distribution models [e.g. @pacifici2017; @guelat2018]. For demographic models, this has rarely been implemented so far, not least due to the fact that few demographic models have sufficient spatial resolution [@schaub2021]. The ptarmigan IDSM presented in this study, however, does have sufficient resolution and our results do indeed support that there is spatial clustering in both overall and time-dependent demographic parameters (e.g. @fig-pVR, SFs "Avg_pSurv_Map.pdf" & "Avg_rRep_Map.pdf"). Furthermore, we did find that mixing of several of the global and area-specific intercept parameters in the current model was suboptimal, suggesting that there may be much to gain from additional structuring, as well as from development of more efficient MCMC sampling strategies for the resulting extended model. One promising framework for approaching this are conditionally autoregressive models [CARs, @verhoef2018]. Such models have been used repeatedly for modelling spatial autocorrelation in species occupancy and demographic rates [e.g. @saracco2010; @saracco2012; @guelat2018] and are straightforward to implement using NIMBLE [@lawson2020]. One possible challenge with using CAR models to explicitly model spatial correlations within our ptarmigan IDSM is that CAR models rely on "neighborhood" relationships between discrete areas and many "neighbors" are missing in our ptarmigan data (e.g. @fig-dens). Estimation of latent parameters in missing areas may be possible though (Perry de Valpine, personal communication; @schaub2021 chapter 19), and this may result in a unique opportunity for making predictions of ptarmigan population trends in unmonitored areas, provided that data for a sufficient number and range of areas are available. Here, we may benefit from the fact that the line transect survey data included in this study constitutes just the publicly available part of the data collected through "Hønsefuglportalen" but the programme also includes additional surveys on private land. Extending to data from private land would provide better coverage especially in south-eastern and southern parts of Norway, which includes areas where only very limited amounts of data are collected on public land. Exploring to what degree additional data from Hønsefuglportalen could be included in future studies employing an extended IDSM with additional spatial structuring is therefore a worthwhile endeavor.
